---
title: "Index.Rmd"
author: "Jonathan Schuster"
date: "08/10/2020"
output: 
  html_document:
    toc: TRUE
    toc_depth: 5  
---
 <!-- This initial chunk is loading all of the packages I'll be utilizing for my COVID report. Example, I'll be using the package "choroplethr" to set up choropleth maps as a form of graph. Additionally, above you'll find I modified my YAML code to add a table of contents wit "toc:TRUE" and allowing it to add any header levels up to 5. Throughout this report you'll find some chunk options that formatted my images, and tables. Example, I used "fig.align=center" to center my image on the html document. --> 

```{r setup, include=FALSE, echo = FALSE}
library(tidyverse)
library(stringr)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(yaml)
knitr::opts_chunk$set(echo = TRUE)
```
### Abstract (tl;dr)
Summer 2020 Ferris State University reproducible research project looking at the Midwest Coalition States COVID death numbers/case numbers compared to states with similar population sizes. 
 
### Introduction
This is a Summer 2020 Ferris State University project for BIOL 390 Reproducible research created by Jonathan Schuster. Data was found from several reputable datasets located from the CDC and used to better understand my hypothesis on the Midwest Coalition's COVID 19 response. R Studio was used to clean datasets, from the CDC, which were used to make visual reputations through graphs/tables on R Markdown.

### Hypothesis
Hypothesis: On April 16, 2020 seven states joined together to form the Midwest Coalition to combat the spead of COVID 19. These states include Michigan, Ohio, Wisconsin, Minnesota, Illinois, Indiana, and Kentucky. My hypothesis is the Midwest Coalition has effectively reduced the spread of COVID since it's creation compared to other states with similar populations such as New Jersey, Georgia, Maryland, Colorado, Pennsylvania, Tennessee, and Louisiana.

 <!-- This chunk is uploading all the databases to my Rmd file. Files are from the CDC. I elected to use two different datasets from the CDC and you'll see below the ways I've elected to manipulate the datasets chosen.   -->
```{r importing, message = FALSE, echo = FALSE}
data(state.regions)
myData <- read_csv("raw_data/US_MAP_DATA.csv", skip = 2)
myData2 <- read_csv("raw_data/Provisional_COVID-19_Death_Counts_by_Place_of_Death_and_State.csv")

```

 <!-- Manipulating data from myData2 from the CDC to show total deaths due to COVID in Midwest Coalition States. I'm selecting 2 columns that help answer my hypothesis, State and COVID19 Deaths, and renaming COVID19 Deaths to a simpler name "Deaths". After renaming, I'm filtering my data to the Midwest coalition states I've selected for my report. Next, I use group_by() function to change my multiple entries of each state into one entry for each state. I use the summarize function to add all the values of Deaths into one number. Lastly, I'm running a function to remove any NA's that exist in my dataset. You'll notice I'm saving this data into a new folder "output" after I've manipulated it. --> 

```{r data2stuff, message=FALSE, echo=FALSE, warning=FALSE}

covid19Data <- myData2 %>% select(`State`, `COVID19 Deaths`) %>% rename(Deaths = `COVID19 Deaths`) %>% group_by(State) %>% filter(State %in% c("Michigan", "Ohio", "Wisconsin", "Minnesota", "Illinois", "Indiana", "Kentucky") & !is.na(`Deaths`)) %>% group_by(State) %>% summarize(Deaths = sum(Deaths))

write.csv(covid19Data,"C:/Users/Jonny/Documents/GitHub/COVID-19/output/Dth_by_state_midwest_coalition_data2.csv", row.names = FALSE)
```

 <!-- Preparing to create a bar plot of death data due to COVID for Midwest Coalition States. I used the States as my x-axis and my total deaths on my y-axis. Geom_col() is telling my ggplot to make a bar graph. Lastly, you'll notice the labs() function and that is to give my graph labels and a title. --> 

```{r ggplotcovid, message=FALSE, echo=FALSE, warning=FALSE, out.width = "500px", out.extra='style="float:right; padding:10px"'}

ggplot(covid19Data, aes(State, Deaths)) + geom_col(color = "black", fill = "indianred2") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(title = "Midwest Coalition States COVID deaths as of Feb-01-2020", x = "State", y = "Deaths") + scale_y_continuous(labels = scales::number_format()) + coord_flip()

```


### Total Death Analysis

On April 16, 2020 seven states came together to "flatten the curve". These states include Michigan, Ohio, Wisconsin, Minnesota, Illinois, Indiana, and Kentucky. They referred this coalition as the Midwest Coalition. On the right you will find two different graphs. On the top (Red), which will contain data on total deaths in states from the Midwest Coalition from Feb 01, 2020. On the bottom (Blue), you will find total death data from seven Non-Coalition states which include New Jersey, Georgia, Maryland, Colorado, Pennsylvania, Tennessee, and Louisiana. 



#### Total Death Midwest Coalition Analysis 

Furthermore, investigating the graphics to the right we can see on average Midwest Coalition states saw fewer overall deaths. The state that saw the least deaths was Kentucky at ~ 700 deaths. The state with the most deaths came out to be Illinois at ~ 6,800 deaths. Surely due to more metropolitan areas. If we were to sum up all the deaths from Midwest Coalition States since February 01, 2020 we would get ~ 21,000 total deaths. 

 <!-- Manipulating data from myData2 from the CDC to show total deaths due to COVID in similar population States. I'm selecting 2 columns that help answer my hypothesis, State and COVID19 Deaths, and renaming COVID19 Deaths to a simpler name "Deaths". After renaming, I'm filtering my data to the similar population states I've selected for my report. Next, I use group_by() function to change my multiple entries of each state into one entry for each state. I use the summarize function to add all the values of Deaths into one number. Lastly, I'm running a function to remove any NA's that exist in my dataset. You'll notice I'm saving this data into a new folder "output" after I've manipulated it. --> 

```{r data2stuff2, message=FALSE, echo=FALSE, warning=FALSE}

covid19Data2 <- myData2 %>% select(`State`, `COVID19 Deaths`) %>% rename(Deaths = `COVID19 Deaths`) %>% filter(State %in% c("New Jersey", "Georgia", "Maryland", "Colorado", "Pennsylvania", "Tennessee", "Louisiana") & !is.na(`Deaths`)) %>% group_by(State) %>% summarize(Deaths = sum(Deaths))

write.csv(covid19Data2,"C:/Users/Jonny/Documents/GitHub/COVID-19/output/Dth_by_state_similar_pop_data2.csv", row.names = FALSE)
```

<!-- Preparing to create a bar plot of death data due to COVID for Similar Population States. I used the States as my x-axis and my total deaths on my y-axis. Geom_col() is telling my ggplot to make a bar graph. Lastly, you'll notice the labs() function and that is to give my graph labels and a title. Very similar to my midwest coalition code. --> 

```{r ggplotcovid2, message=FALSE, echo=FALSE, warning=FALSE, out.width = "500px", out.extra='style="float:right; padding:10px"'}

ggplot(covid19Data2, aes(State, Deaths)) + geom_col(color = "black", fill = "lightblue") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(title = "Non-coalition States COVID deaths as of Feb-01-2020", x = "State", y = "Deaths") + scale_y_continuous(labels = scales::number_format()) + coord_flip()

```

#### Total Death Non-Coalition Analysis

Investigating Non-coalition states we can see slightly more overall deaths. The state that saw the least deaths was Tennessee at ~ 900 deaths. The state with the most deaths came out to be New Jersey at ~ 14,000 deaths. If we were to sum up all the deaths from Non-coalition States since February 01, 2020 we would get ~ 34,000 total deaths. That's over 10,000 deaths more than states residing in the Midwest Coalition.


 <!-- In this chunk I am manipulating a data set into Midwest Coalition states and making a choropleth representation. I'm renaming "Total Cases" to "Cases" for simplicity in writing my code. I've filtered my data set into my 7 coalition states and changed the case to lower case with the str_to_lower function on jurisdiction. Lastly, I'm saving this csv into output folder now that I've manipulated the raw data. --> 

```{r tidyup, message=FALSE, echo=FALSE, warning=FALSE}

caseData <- myData %>% 
        rename(Cases = "Total Cases") %>%
        summarise(region = str_to_lower(jurisdiction), value = Cases) 

write.csv(caseData,"C:/Users/Jonny/Documents/GitHub/COVID-19/output/deaths_by_state_midwest_coalition.csv", row.names = FALSE)

```




### Total Case Analysis 

Up next, I'll be displaying information in the form of a choropleth map on total cases arising from Midwest coalition states, and Non-coalition states, respectively. Down below, you'll find two maps focusing on the coalition and non-coalition states.  

 <!-- In this chunk I'm taking my manipulated data and making my choropleth map for Midwest coalition states. I've given the graph a nice title and labeled my legend "Deaths" to give viewers a way to decipher my image. --> 

```{r choropleth, message=FALSE, echo=FALSE, warning=FALSE, out.width = "500px", fig.align='center'}

state_choropleth(caseData, title = "Total Cases from COVID", legend = "Cases", zoom = c("michigan", "ohio", "wisconsin", "minnesota", "illinois", "indiana", "kentucky"))

```

#### Total Cases Midwest Coalition Analysis

Let's start with the visual colors. The legend tells us the numbers we are looking at describe the total amount of cases each state is seeing. The darker the color represents more cases recorded. The nice part about this graphic is the legend tells us the exact number of cases each state shows. The state in the Midwest coalition with the least reported cases is Kentucky at 34,758 cases. The state showing the most reported cases was Illinois with 193,998 cases. 

 <!-- In this chunk I am manipulating a data set into comparison states and making a choropleth representation. I'm renaming "Total Cases" to "Cases" for simplicity in writing my code just like with my Midwest states. I've filtered my data set into my 7 similar pop states and changed the case to lower case with the str_to_lower function on jurisdiction. Lastly, I'm saving this csv into my output folder now that I've manipulated the raw data. --> 

```{r tidyup2, message=FALSE, echo=FALSE, warning=FALSE}

caseData2 <- myData %>% 
        rename(Cases = "Total Cases") %>%
        summarise(region = str_to_lower(jurisdiction), value = Cases)

write.csv(caseData2,"C:/Users/Jonny/Documents/GitHub/COVID-19/output/deaths_by_state_similar_pop.csv", row.names = FALSE)

```

 <!-- In this chunk I'm taking my manipulated data and making my choropleth map for my selected comparison states. I've given the graph a nice title and labeled my legend "Deaths" to give viewers a way to decipher my image. --> 

```{r choropleth2, message=FALSE, echo=FALSE, warning=FALSE, out.width = "500px", fig.align='center'}

state_choropleth(caseData2, title = "Total Cases from COVID", legend = "Cases", zoom = c("new jersey", "georgia", "maryland", "colorado", "pennsylvania", "tennessee", "louisiana"))

```
 
#### Total Cases Non-Coalition Analysis

Working with the same parameters as above we now see the Non-coalition data on cases. The state with the least reported cases is Colorado at 50,324 cases. The state showing the most reported cases was Georgia with 213,427 cases. Looking at only the most and least recorded states we see how Non-coalition states reported more cases on both ends of the spectrum. 

### Table Analysis

#### Midwest Coalition States
 
 <!-- In this chunk I'm attempting to summarize the data to get an idea about the mean deaths for the Midwest coalition states. Additionally, I'm renaming several columns with dplyr function rename(). After renaming, I elect to filter my results to my chosen 7 states. After filtering, I'm grouping my results by jurisdiction or the 7 states. After tidying up my results I shoot over to kable to create my table. Here I'm selecting the columns I want to present with the select() function. After, I'm adding a table caption and some other neat features like text size. --> 

```{r summary, message=FALSE, echo=FALSE, warning=FALSE, include=TRUE}

summaryData <- myData %>% rename(Cases = "Total Cases", Deaths = "Total Death", Deaths_per_100k = "Death_100k") %>% filter(jurisdiction %in% c("Michigan", "Ohio", "Wisconsin", "Minnesota", "Illinois", "Indiana", "Kentucky")) %>% group_by(jurisdiction)

table1 <- knitr::kable(summaryData %>% select(Cases, Deaths, CasesInLast7Days, Deaths_per_100k),caption = "Table 1: Summary of number of cases/deaths and cases in the last 7 dats for Midwest Coalition States")
kable_styling(table1, position = "right", latex_options = "striped", font_size = 14)
```

#### Non-coalition States


 <!-- In this chunk I'm attempting to summarize the data to get an idea about the mean deaths for the states I've selected to pair against the Midwest coalition states. Additionally, I'm renaming several columns with dplyr function rename(). After renaming, I elect to filter my results to my chosen 7 states. After filtering, I'm grouping my results by jurisdiction or the 7 states. After tidying up my results I shoot over to kable to create my table. Here I'm selecting the columns I want to present with the select() function. After, I'm adding a table caption and some other neat features like text size. --> 

```{r summary2, message=FALSE, echo=FALSE, warning=FALSE, include=TRUE}

summaryData2 <- myData %>% rename(Cases = "Total Cases", Deaths = "Total Death", Deaths_per_100k = "Death_100k") %>% filter(jurisdiction %in% c("New Jersey", "Georgia", "Maryland", "Colorado", "Pennsylvania", "Tennessee", "Louisiana")) %>% group_by(jurisdiction)

table2 <- knitr::kable(summaryData2 %>% select(Cases, Deaths, CasesInLast7Days, Deaths_per_100k), caption = "Table 2: Summary of number of cases/deaths and cases in the last 7 days for Non-coalition States")
kable_styling(table2, position = "right", latex_options = "striped", font_size = 14)
```

### Conclusions

Drawing up some conclusions over the tables and rest of the report. First, let's go over the tables. In the Midwest Coalition states we are looking at two new variables (Cases in the last 7 days, and Deaths per 100,000 people). If we get the mean "Cases in the last 7 days" we get ~ 6,613 new cases in the last 7 days for all seven coalition states combined. Looking at the "Deaths per 100,000" for coalition states we get a mean of 38 deaths per 100,000 people. When looking at the Non-coalition states "Cases in the last 7 days" and "Deaths per 100,000" we get mean values of ~ 7,826, and 68, respectively. So when comparing these results alone we see that Coalition states saw over 1,000 less cases on average compared to states without a coalition. When looking at deaths per 100,000 Coalition states saw over 30 less deaths on average per 100,000 compared to Non-coalition states. Looking back on our choropleths we saw Coalition states contained the state with the least number of reported cases due to COVID and Non-Coalition states contained the state with the highest total reported cases due to COVID. Lastly, when looking at our bar graphs we came up with 10,000 more deaths collectively from our Non-coalition states compared to that of our Midwest Coalition states. 

#### Answering the Hypothesis

Based on all of the results from the bar charts, choropleth maps, and table summaries. The formation of the Midwest Coalition has successfully responded to COVID in a positive way compared to states without a coalition. Coalition states have witnessed less reports of overall cases and less total deaths from the virus COVID 19. 

#### Discussion

Some key points to note in this report. First, when selecting the non-coalition states I only considered total state populations. Each state varied between population differences anywhere from 2,000 - 1,000,000 population differences were used. For example, I used New Jersey compared to Michigan. New Jersey population is 8,958,000 and Michigan has 9,922,000. The difference comes out to 964,000. Additionally, I did not take into consideration population densities throughout each state. For example, being from Michigan I know a majority of deaths/cases come from Detroit or Wayne County. I recognize that metropolitan area's will contain skewed data for each state. That being said this project is at an undergraduate level and I did not narrow my results as a factor. Another point to mention, would be the lack of research on whether or not the selected Non-coalition states actually do reside in a coalition (Just not in the Midwest coalition). Lastly, I am new, or a beginner, to the R language. With that being said, my coding skills held me back from summarizing current COVID data to truly give accurate insights to whether or not the Midwest Coalition truly "flattened the curve" compared to states that did not collaborate ways to slow the progress of the disease.  

### References

[https://www.cdc.gov/covid-data-tracker/#cases](https://www.cdc.gov/covid-data-tracker/#cases) - Used Cases, and Deaths data

[https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Place-of-Deat/uggs-hy5q](https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Place-of-Deat/uggs-hy5q) - Used COVID19 Death data

[https://www.nationsonline.org/oneworld/US-states-population.htm](https://www.nationsonline.org/oneworld/US-states-population.htm) - Used to find seven states with similar population sizes as Midwest Coalition States



